---
title: "Machine Learning Project"
author: "Robert Cordery"
date: "Monday, March 21, 2015"
output: html_document
---

## Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. In this project, the goal is to predict the quality of exercise using data from accelerometers on 6 participants. More information, and their paper is available from the website here: http://groupware.les.inf.puc-rio.br/har. 

```{r setEnvironment}
# Load teh necessary libraries
library(caret)
library(ggplot2)
library(GGally)
library(dplyr)
```

## Load data 

The training data for this project are available here: 

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data has several variables that are all or mostly NA or blank. I remove those from the loaded data. 

```{r loadData, cache =TRUE}

trainFile <- "pml-training.csv"
testFile <- "pml-testing.csv"
dataUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn"
dataLocal <- "./data"
dir.create(dataLocal, showWarnings = FALSE)

trainSource <- paste(dataLocal,trainFile, sep="/")
if(!file.exists(trainSource)){
    download.file(url = paste(dataUrl,trainFile, sep="/"),
                  destfile = trainSource, method="curl")
} 
trainData <- read.csv(trainSource, na.strings=c("NA","","#DIV/0!"))
trainData$classe <- factor(trainData$classe)
trainData$user_name <- factor(trainData$user_name)
useCol <- colSums(is.na(trainData)) < 1000
trainData <- trainData[,useCol]
testSource <- paste(dataLocal,testFile, sep="/")
if(!file.exists(testSource)){
        download.file(url = paste(dataUrl,testFile, sep="/"),
                  destfile = testSource, method="curl")

} 
testData <- read.csv(testSource)
testData <- testData[,useCol]
```

# Partition the data

The trainData set is quite large with a large number of variables, so I decided to split the data rows randomly into three parts:  

1. an initial set of 20% of the data is used to create an initial model to identify the most important variables,
2. a larger training set of %60 of the data is used to train a random forest model,
3. the remaining 20% set is used for cross validation and to estimate the confusion matrix. 

Variables that are not related to motion are removed from the data.  For example, I removed the username variable. While the result may be correlated with username, the goal is to predict the quality of exercise for another person, not for the particular test subjects.  

```{r partitionData}
initialPartition <- createDataPartition(y=trainData$classe,p=0.2, times=2,list=FALSE)
indexInit <- initialPartition[,1]
indexTest <- initialPartition[,2]
indexTrain<- setdiff(1:nrow(trainData),  
                   union(indexInit,indexTest))
# Construct the initial, training and test data sets
initSet <- trainData[indexInit,8:60]
testSet <- trainData[indexTest,8:60]
trainSet<- trainData[indexTrain,8:60]
```

# Select important variables using a rough model

Here we use the initial set to select variables that meet a minimum importance level. 

```{r VariableSelection, cache=TRUE}

modFit <- train(classe ~ ., data=initSet, method="rf", prox=TRUE, ntree=30, tuneGrid = NULL, tuneLength = 3)

minImportance <- 8
importanceVars <- varImp(modFit)
useVar <- importanceVars$importance > minImportance
sum(useVar)

# Add in the output variable classe
useVar=c(useVar,TRUE)
trainSet2<-trainSet[,useVar]
testSet2 <-testSet[,useVar]

```
## Train the final model

The final model is trained on the training set with the reduced set of variables. The selected variables are `r  names(testSet2)`.

```{r fitTraining, cache=TRUE}

fitControl <- trainControl(method="cv" )
modFitTrain <- train(classe ~ ., trControl=fitControl, data=trainSet2, method="rf", prox=TRUE, ntree=40,tuneGrid = NULL, tuneLength = 3)
importanceVarsTrain <- varImp(modFitTrain)

```

## Cross validation

Now we use the part of the data set aside for testing the final model and construct a confusion matrix. 
```{r crossValidation}

predictions <- predict(modFitTrain, newdata = testSet2)

a=c("classe", "roll_belt", "pitch_forearm", "yaw_belt")


ggpairs(trainSet[,a], lower = list(continuous = "smooth"),params = c(method = "loess"))

confusionMatrix(predictions, testSet2$classe)

```
## Apply final model to the testData
The classe variable is not known for the test data set. 
```{r}
testPredictions <- predict(modFitTrain, newdata = testData)
answers=as.character(testPredictions)
answers
```

## Generate submissions
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

answers=as.character(testPredictions)
pml_write_files(answers)

```

## References
Velloso, E.; Bulling, A.; Gellersen, H.; Ugulino, W.; Fuks, H. Qualitative Activity Recognition of Weight Lifting Exercises. Proceedings of 4th Augmented Human (AH) International Conference in cooperation with ACM SIGCHI (Augmented Human'13) . Stuttgart, Germany: ACM SIGCHI, 2013.
